A High Efficiency MAC Protocol for WLANs: Providing Fairness in Dense Scenarios

### 2015-10-19 introProtocol.tex
Grouped images by traffic type. Need more reduction, though

Collisions are handled as in CSMA/CA, which is described in Algorithm~\ref{alg:CSMA_CA}. In Algorithm~\ref{alg:CSMA_CA}, the node starts by setting the retransmissions counter and backoff stage to zero ($r\in[0,R]$ and $k\in[0,m]$ respectively, $m$ is the maximum backoff stage of typical value $m=5$, and $R$ is the retransmissions limit with typical value $R=m+1$), then generates a random backoff, $B$. When the Acknowledgement (\emph{ack}) for a sent packet is not received by the sender a collision is assumed. Upon collision, the involved nodes will double their contention window by incrementing their backoff stage in one and using a random backoff, $B\in[0,2^{k}CW_{\min}]$. This procedure is described between Line~\ref{collision} and~\ref{finalCollision} of Algorithm~\ref{alg:CSMA_CA}.

Algorithm~\ref{alg:CSMA_ECA} provides an explanation of CSMA/ECA deterministic backoff mechanism, which main difference with CSMA/CA (and therefore with Algorithm~\ref{alg:CSMA_CA}) relies on the selection of a deterministic backoff after a successful transmission (compare Line~\ref{randomBackoff} in Algorithm~\ref{alg:CSMA_CA} with Line~\ref{deterministicBackoff} in Algorithm~\ref{alg:CSMA_ECA}). Figure~\ref{fig:BECA-example} shows an example of CSMA/ECA dynamics with four contenders.

@@

Referring to Figure~\ref{fig:BECA}, CSMA/ECA is able to achieve an aggregated throughput that goes beyond CSMA/CA up until the number of contenders ($N$) is greater than $B_{\text{d}}$ ($B_{\text{d}}=7$ in the case of the figure). Beyond this point, the network will have a mixed behavior relating to backoff mechanisms: some nodes will successfully transmit and use a deterministic backoff while others will collide due to the lack of empty slots and return to a random backoff. As more contenders join the network, CSMA/ECA performance will approximate to CSMA/CA's. \textcolor{red}{This effect is also observed in L-ZC and L-MAC, albeit at a larger number of nodes. This is becasue the virtual cicle used by these protocolos ($N$ for L-ZC and $C$ for L-MAC in Section~\ref{relatedWork}) in greater than B$_{d}$. Specifically, the virtual cycle is of size equal to $16$.}

@@

	Figure~\ref{fig:ECA-vs-DCF-maxAgg} shows the aggregated throughput for CSMA/ECA$_{\text{Hys+FS}}$, CSMA/ECA$_{\text{Hys+MaxAg}}$, CSMA/CA using the Fair Share mechanism (CSMA/CA$_{\text{FS}}$), and CSMA/CA$_{\text{MaxAg}}$. Although CSMA/CA$_{\text{FS}}$ and CSMA/CA$_{\text{MaxAg}}$ perform aggregation, collisions degrade the aggregated throughput as more contenders attempt transmission. On the other hand, CSMA/ECA$_{\text{Hys+FS}}$ is able to build a collision-free schedule and takes advantage of the aggregation provided by Fair Share, which opposed to just using CSMA/ECA$_{\text{Hys+MaxAg}}$, it is fair (see Figure~\ref{fig:fairness-with-maximum-aggregation}).

@@

	\begin{figure}[tb]
	\centering
		\includegraphics[width=0.7\linewidth,angle=-90]{figures/fairness-max-Ag/fairness-maxAg-combined-TON.eps}
		\caption{Fairness comparision when using maximum aggregation}
		\label{fig:fairness-with-maximum-aggregation}
	\end{figure}

>>>>

Collisions are handled as in CSMA/CA, which is described in Algorithm~\ref{alg:CSMA_CA}. In Algorithm~\ref{alg:CSMA_CA}, the node starts by setting the retransmissions counter and backoff stage to zero ($r\in[0,R]$ and $k\in[0,m]$ respectively, $m$ is the maximum backoff stage of typical value $m=5$, and $R$ is the retransmissions limit with typical value $R=m+1$), then generates a random backoff, $B$. When the Acknowledgement (\emph{ack}) for a sent packet is not received by the sender a collision is assumed. Upon collision, the involved nodes will double their contention window by incrementing their backoff stage in one and use a random backoff, $B\in[0,2^{k}CW_{\min}]$. This procedure is described between Line~\ref{collision} and~\ref{finalCollision} of Algorithm~\ref{alg:CSMA_CA}.

Algorithm~\ref{alg:CSMA_ECA} provides an explanation of CSMA/ECA's deterministic backoff mechanism, which main difference with CSMA/CA (and therefore with Algorithm~\ref{alg:CSMA_CA}) relies on the selection of a deterministic backoff after a successful transmission (compare Line~\ref{randomBackoff} in Algorithm~\ref{alg:CSMA_CA} with Line~\ref{deterministicBackoff} in Algorithm~\ref{alg:CSMA_ECA}). Figure~\ref{fig:BECA-example} shows an example of CSMA/ECA dynamics with four contenders.

@@

Referring to Figure~\ref{fig:BECA}, CSMA/ECA is able to achieve an aggregated throughput that goes beyond CSMA/CA up until the number of contenders ($N$) is greater than $B_{\text{d}}$ ($B_{\text{d}}=7$ in the case of the figure). Beyond this point, the network will have a mixed behavior relating to backoff mechanisms: some nodes will successfully transmit and use a deterministic backoff while others will collide due to the lack of empty slots and return to a random backoff. As more contenders join the network, CSMA/ECA performance will approximate to CSMA/CA's. \textcolor{red}{This effect is also observed in L-ZC and L-MAC, albeit at a larger number of nodes. This is becasue the virtual cicle used by these protocolos ($N$ for L-ZC and $C$ for L-MAC in Section~\ref{relatedWork}) in greater than B$_{d}$. Specifically, the virtual cycle is of size equal to $16$ (changing to $B_{\text{d}}=16$ would only delay CSMA/ECA's throughput degradation to $N=B_{\text{d}}$).}

@@

	Figure~\ref{fig:ECA-vs-DCF-maxAgg} shows the aggregated throughput for CSMA/ECA$_{\text{Hys+FS}}$, CSMA/ECA$_{\text{Hys+MaxAg}}$, CSMA/CA using the Fair Share mechanism (CSMA/CA$_{\text{FS}}$), and CSMA/CA$_{\text{MaxAg}}$. JFIs are shown below. Although CSMA/CA$_{\text{FS}}$ and CSMA/CA$_{\text{MaxAg}}$ perform aggregation, collisions degrade the aggregated throughput as more contenders attempt transmission. On the other hand, CSMA/ECA$_{\text{Hys+FS}}$ is able to build a collision-free schedule and takes advantage of the aggregation provided by Fair Share, which opposed to just using CSMA/ECA$_{\text{Hys+MaxAg}}$, it is fair.

@@

	%\begin{figure}[tb]
%	\centering
%		\includegraphics[width=0.7\linewidth,angle=-90]{figures/fairness-max-Ag/fairness-maxAg-combined-TON.eps}
%		\caption{Fairness comparision when using maximum aggregation}
%		\label{fig:fairness-with-maximum-aggregation}
%	\end{figure}

### 2015-10-19 journal.tex
Grouped images by traffic type. Need more reduction, though

\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage{float}
\usepackage{color, colortbl}
\usepackage{xcolor}
\usepackage{array}
\usepackage{multirow}
\usepackage{footnote}
\usepackage{cite}
%The below is used to add notes to tables without disrupting the IEEEtran format
\usepackage{threeparttable}
%Multiple figures in a row
%\usepackage{caption}
%\usepackage{subcaption}
\usepackage{amsmath}
\usepackage{mathtools}
%Algorithms
\usepackage[linesnumbered,vlined]{algorithm2e}

@@

This work describes CSMA/ECA and its extensions. Additionally, it provides the first evaluation results of CSMA/ECA in non-saturated traffic conditions as well as its performance when coexisting with CSMA/CA nodes. Furthermore, the effects of imperfect clocks over CSMA/ECA's deterministic backoff mechanism and its consequences when attempting to implement the protocol in real hardware are also analysed.

\end{abstract}

\begin{IEEEkeywords}
CSMA/ECA, WLAN, MAC, Collision-free, Clock Drift.
\end{IEEEkeywords}

@@

Beyond simulations results, the implementation of CSMA/ECA prototypes\cite{ECA-DEMO-INFOCOM14, sanabria2013prototyping, BECA-test,CF-MAC} show that the construction of collision-free schedules using a deterministic backoff after successful transmissions is possible and results in a throughput increase by reducing the number of corrupted frames. \textcolor{red}{We then present the first real hardware implementation of CSMA/ECA with Hysteresis using the open firmware OpenFWWF~\cite{OpenFWWF}. Results show that CSMA/ECA is able of providing higher throughput and lower collisions than CSMA/CA for the same number of contenders.}

An overview of similar distributed and collision-free MAC protocols for WLANs is provided in Section~\ref{relatedWork}. CSMA/ECA, as well as its extensions for allocating many contenders in a collision-free schedule are explained in Section~\ref{introProtocol}. Section~\ref{simulations} details the simulation environment for testing CSMA/ECA, while Section~\ref{results} explains the results. An overview of CSMA/ECA real-hardware implementations is compiled in Section~\ref{EDCA}, followed by a summary of the still missing CSMA/ECA features needed to become the next CSMA/CA replacement in Section~\ref{ECAtoCA}. Conclusions are drawn in Section~\ref{conclusions}.

>>>>

\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{float}
\usepackage{color, colortbl}
\usepackage{xcolor}
\usepackage{array}
\usepackage{multirow}
\usepackage{footnote}
\usepackage{cite}
%The below is used to add notes to tables without disrupting the IEEEtran format
\usepackage{threeparttable}
%Multiple figures in a row
%\usepackage{caption}
%\usepackage{subcaption}
\usepackage{amsmath}
\usepackage{mathtools}
%Algorithms
\usepackage[linesnumbered,vlined]{algorithm2e}

@@

\textcolor{red}{This work describes CSMA/ECA and its extensions. Additionally, it provides the first evaluation results of CSMA/ECA with non-saturated traffic, channel errors, and its performance when coexisting with CSMA/CA nodes. Furthermore, it describes the effects of imperfect clocks over CSMA/ECA and present a mechanism to leverage the impact of channel errors over collision-free schedules. Finally, experimental results on throughput and lost packets from a CSMA/ECA implementation using commercial hardware and open-source firmware are presented.}
\end{abstract}

\begin{IEEEkeywords}
CSMA/ECA, WLAN, MAC, Collision-free, \textcolor{red}{Implementation}.
\end{IEEEkeywords}

@@

Beyond simulations results, the implementation of CSMA/ECA prototypes\cite{ECA-DEMO-INFOCOM14, sanabria2013prototyping, BECA-test,CF-MAC} show that the construction of collision-free schedules using a deterministic backoff after successful transmissions is possible and results in a throughput increase by reducing the number of corrupted frames. \textcolor{red}{We then present the first real hardware implementation of CSMA/ECA with Hysteresis using the open firmware OpenFWWF~\cite{OpenFWWF}. Results show that CSMA/ECA is able of providing higher throughput and lower losses than CSMA/CA for the same number of contenders.}

An overview of similar distributed and collision-free MAC protocols for WLANs is provided in Section~\ref{relatedWork}. CSMA/ECA, as well as its extensions for allocating many contenders in a collision-free schedule are explained in Section~\ref{introProtocol}. Section~\ref{simulations} details the simulation environment for testing CSMA/ECA, while Section~\ref{results} explains the results. An overview of CSMA/ECA real-hardware implementation and testbed description is compiled in Section~\ref{EDCA}, followed by a summary of the still missing CSMA/ECA features needed to become the next CSMA/CA replacement in Section~\ref{ECAtoCA}. Conclusions are drawn in Section~\ref{conclusions}.

### 2015-10-19 realImplementation-Gringoli.tex
Grouped images by traffic type. Need more reduction, though

>>>>

We performed four experiments for every number of contenders, starting from 1 up to 25. For every experiment each station establishes an iPerf~\cite{tirumala2005iperf} session and transmits saturated UDP traffic towards a central AP, which also served as iPerf server for each flow. We used WiFi channel 14  in order to avoid interference from other networks. The aggregated throughput is derived from the iPerf logs, while the percentage of lost frames is reported by the firmware. This is known to be $(\text{tx}-\text{sx})/\text{tx}$; where tx are the number of transmission attempts, and sx are the number of acknowledged frames.

### 2015-10-19 results.tex
Grouped images by traffic type. Need more reduction, though

\section{Results}\label{results}

@@

	\begin{figure}[tb]
	\centering
		\includegraphics[width=0.7\linewidth,angle=-90]{figures/saturated/throughput-saturated-w-BOS/throughput-saturated-w-BOS-TON.eps}
		\caption{Throughput under saturated conditions}
		\label{fig:throughput-sat}
	\end{figure}

	\begin{figure}[tb]
	\centering
		\includegraphics[width=0.7\linewidth,angle=-90]{figures/saturated/collisions-saturated/collisions-saturated-TON.eps}
		\caption{Average percentage of collision slots: the fraction of time slots containing collisions.}
		\label{fig:collisions-sat}
	\end{figure}

	\begin{figure}[tb]
	\centering
		\includegraphics[width=0.7\linewidth,angle=-90]{figures/saturated/slots/Pc-evolution-TON.eps}
		\caption{Evolution of the fraction of collision slots in a scenario with 70 saturated stations.}
		\label{fig:collisions-evolution}
	\end{figure}

@@

	\begin{figure}[tb]
		\centering
		\includegraphics[width=0.7\linewidth,angle=-90]{figures/saturated/timeBetweenSxTx-sat/timeBetweenSxTx-multiplot-sat-TON.eps}
		\caption{Average time between successful transmissions: averaging over all stations}
		\label{fig:serviceTime-sat}
	\end{figure}

@@

	\subsection{Non-saturated nodes}\label{resultsUnsaturated}
	Emptying the MAC queue in CSMA/ECA$_{\text{Hys+FS}}$ means that nodes will reset their backoff stage and use a random backoff when a new packet arrives at the queue, breaking the collision-free schedule (if any) for CSMA/ECA$_{\text{Hys+FS}}$ contenders. The following show the impact over throughput, delay and time between successful transmissions when using CSMA/CA and CSMA/ECA$_{\text{Hys+FS}}$ in non-saturated conditions.\\

@@

   	\begin{figure}[tb]
		\centering
		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/throughput-unsaturated/throughput-unsaturated-w-BOS-TON.eps}
		\caption{Throughput in non-saturation conditions}
		\label{fig:throughputUnsat}
	\end{figure}

@@

   	\begin{figure}[tb]
		\centering
		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/collision-unsaturated/collisions-unsaturated-TON.eps}
		\caption{Average percentage of collision slots: the fraction of time slots containing collisions in non-saturated conditions}
		\label{fig:collisions-unsat}
	\end{figure}

@@

	\begin{figure}[tb]
		\centering
		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/queueingDelay/queueingDelay-TON.eps}
		\caption{Average system delay: averaging over all stations}
		\label{fig:serviceTime-unsat}
	\end{figure}

@@

	\begin{figure}[tb]
		\centering
		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/timeBetweenSxTx-unsat/timeBetweenSxTx-multiplot-unsat-TON.eps}
		\caption{Average time between successful transmissions in non-saturated conditions: averaging over all stations}
		\label{fig:timeBetweenSxTx-multiplot-unsat}
	\end{figure}

>>>>

\section{Results}\label{results}
	\begin{figure*}[tb]
	\centering
		\subcaptionbox{Throughput under saturated conditions\label{fig:throughput-sat}}{\includegraphics[width=0.33\linewidth,angle=-90]{figures/saturated/throughput-saturated-w-BOS/throughput-saturated-w-BOS-TON.eps}}\hfill
		\subcaptionbox{Average percentage of collision slots: the fraction of time slots containing collisions\label{fig:collisions-sat}}{\includegraphics[width=0.33\linewidth,angle=-90]{figures/saturated/collisions-saturated/collisions-saturated-TON.eps}}\\
		\subcaptionbox{Evolution of the fraction of collision slots in a scenario with 70 saturated stations\label{fig:collisions-evolution}}{\includegraphics[width=0.33\linewidth,angle=-90]{figures/saturated/slots/Pc-evolution-TON.eps}}\hfill
		\subcaptionbox{Average time between successful transmissions: averaging over all stations\label{fig:serviceTime-sat}}{\includegraphics[width=0.33\linewidth,angle=-90]{figures/saturated/timeBetweenSxTx-sat/timeBetweenSxTx-multiplot-sat-TON.eps}}\\
		\caption{Saturated traffic results}
		\label{fig:satTest}
	\end{figure*}

@@

%	\begin{figure*}
%		\captionsetup{justification=raggedright}
%		\centering
%		\begin{minipage}[t]{0.45\textwidth}
%			\includegraphics[width=0.7\linewidth,angle=-90]{figures/saturated/throughput-saturated-w-BOS/throughput-saturated-w-BOS-TON.eps}
%			\caption{Throughput under saturated conditions}
%			\label{fig:throughput-sat}
%		\end{minipage}\hfill
%		\begin{minipage}[t]{0.45\textwidth}
%			\includegraphics[width=0.7\linewidth,angle=-90]{figures/saturated/collisions-saturated/collisions-saturated-TON.eps}
%			\caption{Average percentage of collision slots: the fraction of time slots containing collisions}
%			\label{fig:collisions-sat}
%		\end{minipage}\\
%		\begin{minipage}[t]{0.45\textwidth}
%			\includegraphics[width=0.7\linewidth,angle=-90]{figures/saturated/slots/Pc-evolution-TON.eps}
%			\caption{Evolution of the fraction of collision slots in a scenario with 70 saturated stations}
%			\label{fig:collisions-evolution}
%		\end{minipage}\hfill
%		\begin{minipage}[t]{0.45\textwidth}
%			\includegraphics[width=0.7\linewidth,angle=-90]{figures/saturated/timeBetweenSxTx-sat/timeBetweenSxTx-multiplot-sat-TON.eps}
%			\caption{Average time between successful transmissions: averaging over all stations}
%			\label{fig:serviceTime-sat}
%		\end{minipage}\hfill
%	\end{figure*}

%	\begin{figure}[tb]
%	\centering
%		\includegraphics[width=0.7\linewidth,angle=-90]{figures/saturated/throughput-saturated-w-BOS/throughput-saturated-w-BOS-TON.eps}
%		\caption{Throughput under saturated conditions}
%		\label{fig:throughput-sat}
%	\end{figure}
%	
%	\begin{figure}[tb]
%	\centering
%		\includegraphics[width=0.7\linewidth,angle=-90]{figures/saturated/collisions-saturated/collisions-saturated-TON.eps}
%		\caption{Average percentage of collision slots: the fraction of time slots containing collisions.}
%		\label{fig:collisions-sat}
%	\end{figure}
%	
%	\begin{figure}[tb]
%	\centering
%		\includegraphics[width=0.7\linewidth,angle=-90]{figures/saturated/slots/Pc-evolution-TON.eps}
%		\caption{Evolution of the fraction of collision slots in a scenario with 70 saturated stations.}
%		\label{fig:collisions-evolution}
%	\end{figure}

@@

%	\begin{figure}[tb]
%		\centering
%		\includegraphics[width=0.7\linewidth,angle=-90]{figures/saturated/timeBetweenSxTx-sat/timeBetweenSxTx-multiplot-sat-TON.eps}
%		\caption{Average time between successful transmissions: averaging over all stations}
%		\label{fig:serviceTime-sat}
%	\end{figure}

@@

%	################################################################################################################

	\subsection{Non-saturated nodes}\label{resultsUnsaturated}

	\begin{figure*}[tb]
	\centering
		\subcaptionbox{Throughput in non-saturation conditions\label{fig:throughputUnsat}}{\includegraphics[width=0.33\linewidth,angle=-90]{figures/unsaturated/throughput-unsaturated/throughput-unsaturated-w-BOS-TON.eps}}\hfill
		\subcaptionbox{Average percentage of collision slots: the fraction of time slots containing collisions in non-saturated conditions\label{fig:collisions-unsat}}{	\includegraphics[width=0.33\linewidth,angle=-90]{figures/unsaturated/collision-unsaturated/collisions-unsaturated-TON.eps}}\\
		\subcaptionbox{Average system delay: averaging over all stations\label{fig:serviceTime-unsat}}{\includegraphics[width=0.33\linewidth,angle=-90]{figures/unsaturated/queueingDelay/queueingDelay-TON.eps}}\hfill
		\subcaptionbox{Average time between successful transmissions in non-saturated conditions: averaging over all stations\label{fig:timeBetweenSxTx-multiplot-unsat}}{\includegraphics[width=0.33\linewidth,angle=-90]{figures/unsaturated/timeBetweenSxTx-unsat/timeBetweenSxTx-multiplot-unsat-TON.eps}}\\
		\caption{Non-saturated traffic results}
		\label{fig:unsatTest}
	\end{figure*}

	Emptying the MAC queue in CSMA/ECA$_{\text{Hys+FS}}$ means that nodes will reset their backoff stage and use a random backoff when a new packet arrives at the queue, breaking the collision-free schedule (if any) for CSMA/ECA$_{\text{Hys+FS}}$ contenders. The following show the impact over throughput, delay and time between successful transmissions when using CSMA/CA and CSMA/ECA$_{\text{Hys+FS}}$ in non-saturated conditions.\\

@@

%   	\begin{figure}[tb]
%		\centering
%		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/throughput-unsaturated/throughput-unsaturated-w-BOS-TON.eps}
%		\caption{Throughput in non-saturation conditions}
%		\label{fig:throughputUnsat}
%	\end{figure}

@@

%   	\begin{figure}[tb]
%		\centering
%		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/collision-unsaturated/collisions-unsaturated-TON.eps}
%		\caption{Average percentage of collision slots: the fraction of time slots containing collisions in non-saturated conditions}
%		\label{fig:collisions-unsat}
%	\end{figure}

@@

%	\begin{figure}[tb]
%		\centering
%		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/queueingDelay/queueingDelay-TON.eps}
%		\caption{Average system delay: averaging over all stations}
%		\label{fig:serviceTime-unsat}
%	\end{figure}

@@

%	\begin{figure}[tb]
%		\centering
%		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/timeBetweenSxTx-unsat/timeBetweenSxTx-multiplot-unsat-TON.eps}
%		\caption{Average time between successful transmissions in non-saturated conditions: averaging over all stations}
%		\label{fig:timeBetweenSxTx-multiplot-unsat}
%	\end{figure}
